Classical Planning Algorithms ¡ on the Atari Video Games

### 2014-11-19 icaps15-paper/paper.tex
submited version to icaps

\input{games-table}
We tested \IW(1) and \BFS\ over \numgames\
 of the $55$ games considered in
\cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
Entries marked as  ``--'' in the table are   still running at the time of submission. 
Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
Eventually all of  the  $55$ games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by $\IW(1)$, \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores reported for \BRFS\ and \UCT\ are taken from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to \textit{all the algorithms}. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except $\IW(1)$, that due to its aggressive pruning does not always use
the full budget and takes less time per action on average.

@@

Last, \IW(1) and \BFS\  in the lookahead select an action every $5$ frames, while \UCT\  selects 
an action in every frame.  This means that in order to explore a branch 
$300$ frames deep, UCT gets to choose $300$ actions, while \IW($1$) and \BFS\ 
get to choose $60$ actions, both however using the same $300$ frames from the
budget. For this, we  followed the setup  of \BRFS\ in BNVB 
that also selects actions every $5$ frames, matching the behavior
of the emulator that requests an action also every $5$ frames. Since the 
lookahead budget is given by a maximum number of (new) frames, and the time 
is mostly taken by calls to the emulator, this may not be the best choice  
for $\IW(1)$ and \BFS\ that may therefore not be exploiting all the 
options afforded by the budget. We hope to investigate the implications of
this gap in the future.

@@

More interestingly,  \IW(1) outscores \UCT\  in $27$ of the \numgames\ games,
while \BFS\ outscores \UCT\ in $21$.
%\footnote{These numbers can actually
%improve further as 6  games for \IW(1) and 4 games for \BFS\ are still running.}
On the other, \UCT\ does better than $\IW(1)$ and \BFS\ in $15$ and $17$ games respectively.
The relative performance between $\IW(1)$ and \BFS\ makes $\IW(1)$  the
best of the two in $22$ games, and  \BFS\  in $12$. In terms of the number of games 
where an algorithm is the overall winner, $\IW(1)$  domainates the others: 
$\IW(1)$ is the overall best in $23$ games, \BFS\ in $11$ games, and \UCT\ in $18$ games. 
Also, \BRFS\ is best in $2$ games (\Centipede, tied up in \Boxing), while the other three algorithms
are tied in another 2 games (\Pong, \Boxing), and $\IW(1)$ and \BFS\ are tied in \TimePilot.

@@

Likewise, in \Freeway\ and \Berzerk\ both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}, that beats \UCT\ on those games.
%% to say the 2 lines below must add context: .. eg
%%% ``which in *these games* outperforms both \UCT\ and the reinforcement learning methods.''
%% add it if that's the case ..
\emph{Perturb} is a simple algorithm
that selects  a fixed  action with probability $0.95$, and a random
action with probability $0.05$.  For \emph{Perturb}, BNVB do not report 
the average score  but the  best score.  \emph{Perturb} manages to do well in domains where rewards
are deep but can be reached by  repeating the same action.
This is the case of \Freeway, where a chicken has to run across a ten lane
highway filled with traffic. Every time the chicken gets across, there is one
unit of reward. If the chicken is hit by a car, it goes back some lanes. 
In \Freeway, only $12$ out of $18$ possible actions have an effect:
$6$ actions in moving the chicken up one lane (up-right, up-left, up-fire, up-right-fire, up-left-fire), 
$6$ in moving the chicken down one lane (down-right, down-left, down-fire, down-right-fire, down-left-fire), 
and $6$ actions do nothing.  \emph{Perturb}  does well in this domain 
when the selected fixed action is one of the $6$ that moves the chicken up.  As noted 
in Table~\ref{performance} and seen in the provided video, \UCT\ does not manage to take the chicken
across the highway at all. The reason for \UCT\ to perform erratically 
is that the roll-outs do not manage to collect any reward. A sequence of actions 
with at least $48$~\footnote{One needs to move up the chicken up for at least $4$ seconds ($240$ frames) in order to get across the 
highway.} actions embedded that move the chicken up are required in order 
to collect a reward. Since the probability for the roll-out to choose an action that moves the chicken up 
is $1/3$, with the parameters chosen by BNVB, the likelihood of \UCT\ generating such a sequence of actions 
that \emph{also} avoids incoming traffic in a timely fashion is rather low.
%% the previous number was not correct; no-op actions do not count, and thus needs
%%% 200 more move up actions than move down actions, although this is not accurate either
%%%% as move down actions that take chicken below the screen count as no-ops too ..
$\IW(1)$ does not have this limitation and is best in \Freeway.

@@

$\IW(1)$ also outperforms the best learning algorithm in
\cite{deep-mind-atari} in the $7$ games considered there, and
\BFS\ does so in $6$ of the $7$ games. Comparing 
with the scores reported for the best Reinforcement Learning algorithm
in \cite{bellemare:jair2013}, we note that either $\IW(1)$ and \BFS\ 
are outperforming significantly the best learning algorithm, in those games
where it is reported to perform better than \UCT\ (\Montezuma, \Venture\
and \Bowling). In \Montezuma\ rewards are very
sparse, deep, and most of the actions lead to losing a life with
no immediate penalty or consequence.  All algorithms achieve $0$ score, 
except for \BFS\ that achieves an average score of $80$, and a score
of $400$ in one of the runs. This means that even \BFS\  is not able to 
consistently find rewards in this game. This game and several other games 
like \Breakout, and \SpaceInvaders\ could be much simpler by adding
negative rewards for losing a life. We have indeed observed that
our planning algorithms do not care much about losing lives
until there is just one life left, when their play noticeably improves.
This can be seen in the videos mentioned above, and suggests that
 some simple form of learning would be useful both
for planners and reinforcement learning approaches.

@@

The notions of width and the \IW\  algorithm guarantee that  states with low width will be generated
in low polynomial time through \emph{shortest} paths. In the presence of rewards like the Atari games,
however, the  interest is not  in the shortest paths but in the \emph{best} paths; i.e, the paths with
maximum reward.  \IW\ may actually  fail to find such paths even when calling \IW($k$) with a high 
$k$ parameter. Optimality  could be  achieved then  by replacing the breadth-first search
underlying \IW($k$) by Dijkstra's algorithm yet such a move would make the relation between
\IW\ and the notion of width less clear. A better  option is to change  \IW\ to comply with 
a different property; namely, that if there is a ``rewarding'' path made up of states of 
low width, then \IW\ will find such paths or better ones in time that is exponential 
in their width. For this,  a simple change in \IW\ suffices: when generating a state 
$s$ that is a \emph{duplicate} of a state $s'$ that has been previously generated and not pruned, 
$s'$ is replaced by $s$ if $R(s) > R(s')$. Moreover in such a case, the change in reward 
must be propagated to the descendants of $R(s')$ that are in memory. This is similar to the change required 
in the  A* search algorithm for preserving optimality when moving from consistent to inconsistent heuristics \cite{pearl:heuristics}. 
The alternative is to ``reopen'' such nodes. The same change is actually needed in \BFS\ to ensure that, 
if given enough time, \BFS\ will actually find optimal paths.
% Changing \BFS\ to account for such paths guarantees that, if given enough time, \BFS\ will actually find
% the optimal paths. 
The code used for \IW\ and \BFS\ in the
experiments above does not implement this change  as the overhead involved in checking for duplicates 
in some test cases did not appear to pay off. More experiments  however are  needed to
find out if this is actually the most effective option.

@@

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\
that originate in work in classical planning,  can be used to play  the Atari video
games where they  achieve  state-of-the-art performance. This level of play 
is the result of a structured exploration of the state space that manages to combine 
the scope of blind search algorithms with the efficiency of heuristic search methods. 
There are several  issues worth studying, we  mention two of them.
Regarding planning, the results show the potential of width-based algorithms
for a broader class of planning problems. As shown, indeed, these algorithms
do not require PDDL encodings and can deal with state successor functions
encoded \emph{procedurally} (as in simulators) as long as the structure of the states 
is known. This suggests to us that the algorithms discussed in the 
present paper could be applied in a broad variety of settings where only on-line 
action selection is feasible or suitable.

Regarding the Atari games, we identify two problems in all the planning methods
discussed. First, by using the RAM state of the Atari console one can consider
that they are ``cheating'', as they use information that is not readily available to the human player, 
and second, they do not get better with experience. In the future, we would like to explore 
whether variations of these algorithms  can be used to play from the information conveyed by the  
 state of screen pixels as opposed to the RAM state, as well as exploring the potential uses of the novel
ideas presented for learning, in order to improve performance.

>>>>

\input{games-table}
We tested \IW(1) and \BFS\ over \numgames\
 of the $55$ games considered in
\cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
Entries marked as  ``--'' in the table are   still running at the time of submission. 
Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
Eventually all of  the  $55$ games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by $\IW(1)$, \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores reported for \BRFS\ and \UCT\ are taken from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to $\IW(1)$, \BFS, and \UCT. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except $\IW(1)$, that due to its aggressive pruning does not always use
the full budget and takes less time per action on average.

@@

Last, \IW(1) and \BFS\  in the lookahead select an action every $5$ frames, while \UCT\  selects 
an action in every frame.  This means that in order to explore a branch 
$300$ frames deep, UCT gets to choose $300$ actions, while \IW($1$) and \BFS\ 
get to choose $60$ actions, both however using the same $300$ frames from the
budget. For this, we  followed the setup  of \BRFS\ in BNVB 
that also selects actions every $5$ frames, matching the behavior
of the emulator that requests an action also every $5$ frames. Since the 
lookahead budget is given by a maximum number of (new) frames, and the time 
is mostly taken by calls to the emulator, this may not be the best choice  
for $\IW(1)$ and \BFS\ that may therefore not be exploiting all the 
options afforded by the budget. We'll look at this further in the future.

@@

More interestingly,  \IW(1) outscores \UCT\  in $27$ of the \numgames\ games,
while \BFS\ outscores \UCT\ in $21$.
%\footnote{These numbers can actually
%improve further as 6  games for \IW(1) and 4 games for \BFS\ are still running.}
On the other hand, \UCT\ does better than $\IW(1)$ and \BFS\ in $15$ and $17$ games respectively.
The relative performance between $\IW(1)$ and \BFS\ makes $\IW(1)$  the
best of the two in $22$ games, and  \BFS\  in $12$. In terms of the number of games 
where an algorithm is the overall winner, $\IW(1)$  domainates the others: 
$\IW(1)$ is the overall best in $23$ games, \BFS\ in $11$ games, and \UCT\ in $18$ games. 
Also, \BRFS\ is best in $2$ games (\Centipede, tied up in \Boxing), while the other three algorithms
are tied in another 2 games (\Pong, \Boxing), and $\IW(1)$ and \BFS\ are tied in \TimePilot.

@@

Likewise, in \Freeway\ and \Berzerk\ both $\IW(1)$ and \BFS\ attain a better
score than the baseline semi-random algorithm \emph{Perturb} in
\cite{bellemare:jair2013}, that beats \UCT\ on those games.
%% to say the 2 lines below must add context: .. eg
%%% ``which in *these games* outperforms both \UCT\ and the reinforcement learning methods.''
%% add it if that's the case ..
\emph{Perturb} is a simple algorithm
that selects  a fixed  action with probability $0.95$, and a random
action with probability $0.05$.  For \emph{Perturb}, BNVB do not report 
the average score  but the  best score.  \emph{Perturb} manages to do well in domains where rewards
are deep but can be reached by  repeating the same action.
This is the case of \Freeway, where a chicken has to run to the top of the screen across a ten lane
highway filled with traffic. Every time the chicken gets across (starting at the bottom), there is one
unit of reward. If the chicken is hit by a car, it goes back some lanes. 
In \Freeway, only $12$ out of the $18$ possible actions have an effect:
$6$ actions move the chicken up (up-right, up-left, up-fire, up-right-fire, up-left-fire), 
$6$ actions move the chicken down (down-right, down-left, down-fire, down-right-fire, down-left-fire), 
and $6$ actions do nothing.  \emph{Perturb}  does well in this domain 
when the selected fixed action moves the chicken up.  As noted 
in Table~\ref{performance} and seen in the provided video, \UCT\ does not manage to take the chicken
across the highway at all. 
%
The reason that \UCT\ does not collect any reward is that it needs to move the chicken up
at least 240 times~\footnote{One needs to move the chicken up for at least $4$ seconds ($240$ frames) in order to get across the 
highway.} something that is very unlikely in a random exploration. $\IW(1)$ does not have this limitation and is best in \Freeway.
%% The reason for \UCT\ to perform erratically 
%% is that the roll-outs do not manage to collect any reward. A sequence of actions 
%% with at least $48$ actions embedded that move the chicken up are required in order 
%% to collect a reward. Since the probability for the roll-out to choose an action that moves the chicken up 
%% is $1/3$, with the parameters chosen by BNVB, the likelihood of \UCT\ generating such a sequence of actions 
%% that \emph{also} avoids incoming traffic in a timely fashion is rather low.
%% %% the previous number was not correct; no-op actions do not count, and thus needs
%% %%% 200 more move up actions than move down actions, although this is not accurate either
%% %%%% as move down actions that take chicken below the screen count as no-ops too ..
%% $\IW(1)$ does not have this limitation and is best in \Freeway.

@@

$\IW(1)$ also outperforms the best learning algorithm in
\cite{deep-mind-atari} in the $7$ games considered there, and
\BFS\ does so in $6$ of the $7$ games. Comparing with the scores reported for the reinforcement learning algorithms by BNVB,
we note that both $IW(1)$ and \BFS\ do much better than the best learning algorithm in
those games where the learning algorithms outperform \UCT\ like \Montezuma, \Venture\
and \Bowling.  In \Montezuma\ rewards are very
sparse, deep, and most of the actions lead to losing a life with
no immediate penalty or consequence.  All algorithms achieve $0$ score, 
except for \BFS\ that achieves an average score of $80$, and a score
of $400$ in one of the runs. This means that even \BFS\  is not able to 
consistently find rewards in this game. This game and several other games 
like \Breakout, and \SpaceInvaders\ could be much simpler by adding
negative rewards for losing a life. We have indeed observed that
our planning algorithms do not care much about losing lives
until there is just one life left, when their play noticeably improves.
This can be seen in the videos mentioned above, and suggest a simple form of learning that would be useful to both planners
and reinforcement learning algorithms.

@@

The notions of width and the \IW\  algorithm guarantee that  states with low width will be generated
in low polynomial time through \emph{shortest} paths. In the presence of rewards like the Atari games,
however, the  interest is not  in the shortest paths but in the \emph{best} paths; i.e, the paths with
maximum reward.  \IW\ may actually  fail to find such paths even when calling \IW($k$) with a high 
$k$ parameter. Optimality  could be  achieved by replacing the breadth-first search
underlying \IW($k$) by Dijkstra's algorithm yet such a move would make the relation between
\IW\ and the notion of width less clear. A better  option is to change  \IW\ to comply with 
a different property; namely, that if there is a ``rewarding'' path made up of states of 
low width, then \IW\ will find such paths or better ones in time that is exponential 
in their width. For this,  a simple change in \IW\ suffices: when generating a state 
$s$ that is a \emph{duplicate} of a state $s'$ that has been previously generated and not pruned, 
$s'$ is replaced by $s$ if $R(s) > R(s')$, with the change of reward propagated to the descendants of $s'$
that are in memory. This is similar to the change required 
in the  A* search algorithm for preserving optimality when moving from consistent to inconsistent heuristics \cite{pearl:heuristics}. 
The alternative is to ``reopen'' such nodes. The same change is actually needed in \BFS\ to ensure that, 
if given enough time, \BFS\ will actually find optimal paths.
% Changing \BFS\ to account for such paths guarantees that, if given enough time, \BFS\ will actually find
% the optimal paths. 
The code used for \IW\ and \BFS\ in the
experiments above does not implement this change  as the overhead involved in checking for duplicates 
in some test cases did not appear to pay off. More experiments  however are  needed to
find out if this is actually the most effective option.

@@

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\
that originate in work in classical planning,  can be used to play  the Atari video
games where they  achieve  state-of-the-art performance. This level of play 
is the result of a structured exploration of the state space that manages to combine 
the scope of blind search algorithms with the efficiency of heuristic search methods. 
There are several  issues worth studying, we  mention two of them.
Regarding planning, the results show the potential of width-based algorithms
for a broader class of planning problems. As shown, indeed, these algorithms
do not require PDDL encodings and can deal with state successor functions
encoded \emph{procedurally} (as in simulators) as long as the structure of the states 
is known. 
%% This suggests to us that the algorithms discussed in the 
%% present paper could be applied in a broad variety of settings where only on-line 
%% action selection is feasible or suitable.

Regarding the Atari games, we identify two problems in all the planning methods
discussed. First, by using the RAM state of the Atari console one can consider
that they are ``cheating'', as they use information that is not available to the human player. And second, they do not get better with experience. In the future, we would like to explore 
whether variations of these algorithms  can be used to play from the information conveyed by the  
 state of screen pixels as opposed to the RAM state, and the potential uses of the new ideas for learning to improve performance.
